Demographic inference using the SFS with \moments and \demes

### 2025-03-31 chapter/paper.tex
small revisions to chapter

\begin{abstract}
    Placeholder
\end{abstract}

@@

The genetic composition of a sample of individuals is shaped by their genome
biology and evolutionary history. Variation resulting from this history can be
fully represented by the ancestral relationships among samples at each locus in
the genome and how those gene-genealogical relationships change along a
chromosome due to recombination (that is, information stored in the Ancestral
Recombination Graph \citep{nielsen2025inference}). However, the ARG can be
large and unwieldy, and methods for reconstructing history directly from the
ARG, while showing promise \citep[e.g.,][]{yc2022evaluation, fan2023likelihood,
brandt2024promise}, are in their infancy and so far limited in application and
scalability. Instead, evolutionary inference using informative summaries of
genetic variation remains a tractable and powerful alternative for learning
parameters of population history, natural selection and genome biology.

@@

Specifying demographic models requires defining populations (or ``demes'') and
their relationships via splits and gene flow. \demes provides a standardized
and accessible format for defining demographic models \citep{gower2022demes},
and has been adopted by widely used software for population genetics simulation
and inference (including \msprime \citep{baumdicker2022efficient}, \momi
\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, \texttt{GADMA}
\citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
code across software, and ease of model implementation and reduction of errors.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using the both the original and misspecified simpler
models. In the second example, we infer a relatively simple human-Neanderthal
history using data from two human populations and a high-coverage ancient
sample from the Neanderthal lineage. For both of these examples, we briefly
describe data processing and inference setup, highlighting the high-level
components of the analyses. For detailed information for each example,
including managing data, specifying models, performing inference, computing
confidence intervals and visualizing fits, we refer readers to the accompanying
GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to the value(s) in the
demographic model and sets lower and upper bounds. This file also specifies any
relative \texttt{constraints} between pairs of parameters, to ensure the
optimization routine only explores valid model space. Below, we have included
a truncated parameters file -- the full example is available on GitHub.
\begin{minted}[gobble=4, frame=single]{YAML}
    parameters:
    - name: T0
      values:
      - demes:
          ancestral:
            epochs:
              0: end_time
      lower_bound: 0
      upper_bound: 2e5
    - name: T1
      values:
      - demes:
          ancestral:
            epochs:
              1: end_time
      lower_bound: 0
      upper_bound: 1e6
    - name: Ne
      values:
      - demes:
          ancestral:
            epochs:
              0: start_size
      lower_bound: 1e2
      upper_bound: 1e6
    ...
    constraints:
    - params: [T0, T1]
      constraint: greater_than
\end{minted}

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
function.  The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.
\begin{minted}[gobble=4, frame=single]{YAML}
    import moments

@@

Confidence intervals may also be computed, once a (locally) optimal model has
been found.

@@

We obtained modern human genome sequences from a recent resequencing of the 1000 Genomes Project cohort \citep{byrska2022high}, and a high-coverage sequence of the Vindija Neandertal from \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR (British from England and Scotland) sequences as our modern human samples.  
After filtering out sites that lay outside the 1000 Genomes `strict' mask, that were not genotyped in all four existing high-coverage archaic genome sequences, lacked high-confidence ancestral state assignments, or fell within 10 kb of exonic or promoter regions, we were left with approximately 960 Mb of well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija sequence to a more recent genome build) can be found in the linked GitHub repository.

The model that we wished to fit is relatively simple but still has many parameters (11 in the example we present), so we constructed it in stages by first fitting simpler one- and two-population graphs.
Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per generation and 29 years as the average generation time.
We began by fitting one-population models with ancestral and recent size changes for MSL and GBR.
For the MSL model, we allowed three epochs with constant effective size in each epoch.
We attributed to GBR a sharp contraction $\sim$60 kya followed by exponential growth to represent the out-of-Africa bottleneck and subsequent rapid expansion. YAML files representing the models and parameters in this section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the resulting graph to get a simple two-population out-of-Africa model that included continuous symmetric migration between MSL and GBR and an ancestral size expansion. 
Following this, we added a Neandertal branch and fitted the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixture proportion. 
We also fitted the divergence time between the sampled Vindija Neandertal and the Neandertal lineage which admixed with modern humans.
Finding that this parameter tended to converge to its lower bound (imposed by the estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature (90 kya, \cite{prufer2017high}) in the second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya. Fixing parameters for which good estimates are available in the literature (see e.g. \cite{sumer2025earliest}) is often a reasonable thing to do in the first stages of model optimization, and we expected this one to be especially difficult to fit due to the small pulse proportion and Neandertal sample size.
In a second round of optimization on the three-population model, we refit all the parameters aside from the two times mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with prior authors, with a Neandertal-modern human divergence time of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 percent. 
We observe some large residuals in our best-fit model (Figure 2), suggesting ...

>>>>

\begin{abstract}
In this chapter we show how to use \moments, a Python library for computing the
joint SFS and inferring demographic history, through its \demes interface. 
Following a discussion of SFS inference, we illustrate the general usage of 
\moments and sketch two specific examples. We conclude with a discussion of some 
additional considerations regarding the \moments library.
\end{abstract}

@@

The genetic composition of a sample of individuals is shaped by their genome
biology and evolutionary history. Variation resulting from this history can be
fully represented by the ancestral relationships among samples at each locus in
the genome and how those gene-genealogical relationships change along a
chromosome due to recombination (that is, by information stored in the Ancestral
Recombination Graph \citep{nielsen2025inference}). However, the ARG can be
large and unwieldy, and methods for reconstructing history directly from the
ARG, while showing promise \citep[e.g.,][]{yc2022evaluation, fan2023likelihood,
brandt2024promise}, are in their infancy and so far limited in application and
scalability. Instead, evolutionary inference using informative summaries of
genetic variation remains a tractable and powerful alternative for learning
parameters of population history, natural selection and genome biology.

@@

Specifying demographic models requires defining populations (or ``demes'') and
their relationships via splits and gene flow. \demes provides a standardized
and accessible format for defining demographic models \citep{gower2022demes},
and has been adopted by widely used software for population genetics simulation
and inference (including \msprime \citep{baumdicker2022efficient}, \momi
\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, \texttt{GADMA}
\citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
code across software, ease of model implementation and reduction of errors.

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using the both the original model and simpler misspecified 
ones. In the second example, we infer a relatively simple human-Neanderthal
history using data from two human populations and a high-coverage ancient
sample from the Neanderthal lineage. For both of these examples, we briefly
describe data processing and inference setup, emphasizing the high-level
components of the analyses. For detailed information about each example,
including managing data, specifying models, performing inference, computing
confidence intervals and visualizing fits, we refer readers to the accompanying
GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

@@

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to desired value(s) in the
demographic model and sets lower and upper bounds. This file also specifies any
relative \texttt{constraints} between pairs of parameters, to ensure the
optimization routine only explores valid model space. Below, we have included
a truncated parameters file -- the full example is available on GitHub.
\begin{minted}[gobble=4, frame=single]{YAML}
    parameters:
    - name: T0
      values:
      - demes:
          ancestral:
            epochs:
              0: end_time
      lower_bound: 0
      upper_bound: 2e5
    - name: T1
      values:
      - demes:
          ancestral:
            epochs:
              1: end_time
      lower_bound: 0
      upper_bound: 1e6
    - name: Ne
      values:
      - demes:
          ancestral:
            epochs:
              0: start_size
      lower_bound: 1e2
      upper_bound: 1e6
    ...
    constraints:
    - params: [T0, T1]
      constraint: greater_than
\end{minted}

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
function.  The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.
\begin{minted}[gobble=4, frame=single]{python}
    import moments

@@

Confidence intervals may also be computed using 
\texttt{moments.Demes.Inference.uncerts(...)}, once a (locally) optimal model 
has been found. The following will emit a tab-separated file containing
parameter names, best-fit values and estimated standard errors.
\begin{minted}[gobble=4, frame=single]{python}
    log_file = 'log.txt'

    uncerts = moments.Demes.Inference.uncerts(
      output, params_file, data,
      uL=U, output=log_file
    )
\end{minted}

@@

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \citep{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal from \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in all four existing high-coverage archaic genome sequences, 
thatlacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions and were left with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many 
parameters (11 in the example we present), so we constructed it in stages by 
first fitting simpler one-- and two--population models.
Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per 
generation and 29 years as the average generation time. We began by fitting 
one-population models with ancestral and recent size changes for MSL and GBR.
For the MSL model, we allowed three epochs with constant effective size in each 
epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
exponential growth to represent the out-of-Africa bottleneck and subsequent 
rapid expansion. YAML files representing the models and parameters in this 
section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the 
resulting graph to get a two-population model that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
(90 kya, \cite{prufer2017high}) in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
(see e.g. \cite{sumer2025earliest}) may be reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
prior authors, with a Neandertal-modern human divergence time of 547 kya, 
MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
percent. We observe some large residuals in our best-fit model 
Figure~\ref{fig:humans}G-I), suggesting ...
